Parallel Tractability of Ontology\\ Materialization: Technique and Practice

### 2018-02-22 parallel-tractability-J-v9/optimizedAlgo.tex
unified the subscripts with \text{}

In previous sections, Algorithm~$\mathsf{A}_{\text{opt}}$ is mainly used for theoretical analysis.
However, this algorithm can hardly work in practice due to its inherently
high requirement of computing space. Specifically,
Algorithm~$\mathsf{A}_{\text{opt}}$ constructs a materialization graph of the given
datalog program $P$ by checking all possible rule
instantiations in $P^*$.
One can check that $|P^*|$ could be the quadratic or cubic in the number of constants occurring in $P$.
Recall the analysis of the parallel complexity of Algorithm~$\mathsf{A}_{bsc}$ and Algorithm~$\mathsf{A}_{\text{opt}}$.
The number of constants is at most $|\textbf{I}|e$ where $e$ denotes the maximum arity of any predicate in $\textbf{I}$.
Consider a datalog rule of the form `$R(x,y),A(y)\rightarrow B(x)$'.
Since there are two variables in this rule, the number of all possible rule instantiations
is $(|\textbf{I}|e)^2$.
Similarly, for a datalog rule rewritten from an axiom of the form (R3) or (R4),
the number of all possible rule instantiations is $(|\textbf{I}|e)^3$.
Undoubtedly, a plain implementation of Algorithm~$\mathsf{A}_{\text{opt}}$ would be slow
when the target datalog program or ontology
tends to be large in size.
On the other hand, from Examples~\ref{exp:mg}, \ref{exp:dllite} and \ref{exp:dhl},
we can observe that the rule instantiations used for the construction
of materialization graphs always cover a small part of $P^*$ with respect to the target datalog program $P$.
Thus, we consider reducing the computing space of Algorithm~$\mathsf{A}_{\text{opt}}$
by narrowing down the scope of rule instantiations to be checked.
Our strategy is to restrict that, \emph{in each iteration of Algorithm~$\mathsf{A}_{\text{opt}}$,
each of the checked rule instantiations should involve at least one body atom that has been
added to the constructed materialization graph}.
Since DHL$(\circ)$
is our focus, we explain how to apply the above strategy in DHL$(\circ)$ materialization
as follows.

We first consider a datalog rule of the
form `$A(x)\rightarrow B(x)$' that corresponds to (T1). The above strategy requires that,
in each iteration, Algorithm~$\mathsf{A}_{\text{opt}}$ can only check the rule instantiations
of the form `$A(a)\rightarrow B(a)$' where $A(a)$ has been added to the constructed materialization
graph $\mathcal{G}$. If there are $n$ atoms of the form $A(x)$ that have been added to $\mathcal{G}$,
the number of all checked rule instantiations of the above rule is also $n$, instead of $(|\textbf{I}|e)$.
This is because, for each assertion $A(a)$,
the datalog rule `$A(x)\rightarrow B(x)$'
has only one rule instantiation `$A(a)\rightarrow B(a)$' where
the variable $x$ is substituted by the constant $a$.
The cases of (R1) and (R2) can be analyzed similarly.
We now consider datalog rules that have two body atoms, i.e.,
datalog rules of the form `$R(x,y),A(y)\rightarrow B(x)$' (see (T3)),
`$A_1(x),A_2(x)\rightarrow B(x)$' (see (T2)) and `$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$' (see (R3) and (R4)).
We require that, for each rule instantiation of these rules checked by Algorithm~$\mathsf{A}_{\text{opt}}$,
at least one body atom has been added to the constructed materialization
graph $\mathcal{G}$; thus, it can be checked that, in each iteration of Algorithm~$\mathsf{A}_{\text{opt}}$,
the number of checked rule instantiations
is at most $k(|\textbf{I}|e)$ where $k$ is the number of atoms that have been
added to $\mathcal{G}$.
Note that, for rule instantiations of two body atoms,
we do not require that both of these two body atoms have been added to
the constructed materialization graph.
Otherwise, the algorithm would
perform as Algorithm~$\mathsf{A}_{bsc}$ and, thus, the
optimizations used in Algorithm~$\mathsf{A}_{\text{opt}}$ cannot further work.

>>>>

In previous sections, Algorithm~$\mathsf{A}_{\text{opt}}$ is mainly used for theoretical analysis.
However, this algorithm can hardly work in practice due to its inherently
high requirement of computing space. Specifically,
Algorithm~$\mathsf{A}_{\text{opt}}$ constructs a materialization graph of the given
datalog program $P$ by checking all possible rule
instantiations in $P^*$.
One can check that $|P^*|$ could be the quadratic or cubic in the number of constants occurring in $P$.
Recall the analysis of the parallel complexity of Algorithm~$\mathsf{A}_{\text{bsc}}$ and Algorithm~$\mathsf{A}_{\text{opt}}$.
The number of constants is at most $|\textbf{I}|e$ where $e$ denotes the maximum arity of any predicate in $\textbf{I}$.
Consider a datalog rule of the form `$R(x,y),A(y)\rightarrow B(x)$'.
Since there are two variables in this rule, the number of all possible rule instantiations
is $(|\textbf{I}|e)^2$.
Similarly, for a datalog rule rewritten from an axiom of the form (R3) or (R4),
the number of all possible rule instantiations is $(|\textbf{I}|e)^3$.
Undoubtedly, a plain implementation of Algorithm~$\mathsf{A}_{\text{opt}}$ would be slow
when the target datalog program or ontology
tends to be large in size.
On the other hand, from Examples~\ref{exp:mg}, \ref{exp:dllite} and \ref{exp:dhl},
we can observe that the rule instantiations used for the construction
of materialization graphs always cover a small part of $P^*$ with respect to the target datalog program $P$.
Thus, we consider reducing the computing space of Algorithm~$\mathsf{A}_{\text{opt}}$
by narrowing down the scope of rule instantiations to be checked.
Our strategy is to restrict that, \emph{in each iteration of Algorithm~$\mathsf{A}_{\text{opt}}$,
each of the checked rule instantiations should involve at least one body atom that has been
added to the constructed materialization graph}.
Since DHL$(\circ)$
is our focus, we explain how to apply the above strategy in DHL$(\circ)$ materialization
as follows.

We first consider a datalog rule of the
form `$A(x)\rightarrow B(x)$' that corresponds to (T1). The above strategy requires that,
in each iteration, Algorithm~$\mathsf{A}_{\text{opt}}$ can only check the rule instantiations
of the form `$A(a)\rightarrow B(a)$' where $A(a)$ has been added to the constructed materialization
graph $\mathcal{G}$. If there are $n$ atoms of the form $A(x)$ that have been added to $\mathcal{G}$,
the number of all checked rule instantiations of the above rule is also $n$, instead of $(|\textbf{I}|e)$.
This is because, for each assertion $A(a)$,
the datalog rule `$A(x)\rightarrow B(x)$'
has only one rule instantiation `$A(a)\rightarrow B(a)$' where
the variable $x$ is substituted by the constant $a$.
The cases of (R1) and (R2) can be analyzed similarly.
We now consider datalog rules that have two body atoms, i.e.,
datalog rules of the form `$R(x,y),A(y)\rightarrow B(x)$' (see (T3)),
`$A_1(x),A_2(x)\rightarrow B(x)$' (see (T2)) and `$R_1(x,y),R_2(y,z)\rightarrow R(x,z)$' (see (R3) and (R4)).
We require that, for each rule instantiation of these rules checked by Algorithm~$\mathsf{A}_{\text{opt}}$,
at least one body atom has been added to the constructed materialization
graph $\mathcal{G}$; thus, it can be checked that, in each iteration of Algorithm~$\mathsf{A}_{\text{opt}}$,
the number of checked rule instantiations
is at most $k(|\textbf{I}|e)$ where $k$ is the number of atoms that have been
added to $\mathcal{G}$.
Note that, for rule instantiations of two body atoms,
we do not require that both of these two body atoms have been added to
the constructed materialization graph.
Otherwise, the algorithm would
perform as Algorithm~$\mathsf{A}_{\text{bsc}}$ and, thus, the
optimizations used in Algorithm~$\mathsf{A}_{\text{opt}}$ cannot further work.

### 2018-02-22 parallel-tractability-J-v9/relatedWork.tex
unified the subscripts with \text{}

The parallel reasoner RDFox \cite{MotikNPHO14} is used to evaluate our implementation. RDFox is a state-of-the-art system that handles reasoning on datalog rewritable ontology languages. Algorithm~$\mathsf{A}_{bsc}$ proposed in Section~\ref{sec:ptclass} is similar to the main algorithm of RDFox (see \cite{MotikNPHO14}, Sections~3 and 4). The difference lies in that, a group of rule instantiations are handled by one processor (namely a thread) in RDFox, while in Algorithm~$\mathsf{A}_{bsc}$, each rule instantiation is assigned to a unique processor. Analogous to the method used in RDFox, ParallelDHL is also implemented by assigning a group of 
rule instantiations to a processor.
The experimental results show that ParallelDHL is comparable to RDFox when handling
the LUBM and the YAGO ontologies.

>>>>

The parallel reasoner RDFox \cite{MotikNPHO14} is used to evaluate our implementation. RDFox is a state-of-the-art system that handles reasoning on datalog rewritable ontology languages. Algorithm~$\mathsf{A}_{\text{bsc}}$ proposed in Section~\ref{sec:ptclass} is similar to the main algorithm of RDFox (see \cite{MotikNPHO14}, Sections~3 and 4). The difference lies in that, a group of rule instantiations are handled by one processor (namely a thread) in RDFox, while in Algorithm~$\mathsf{A}_{\text{bsc}}$, each rule instantiation is assigned to a unique processor. Analogous to the method used in RDFox, ParallelDHL is also implemented by assigning a group of
rule instantiations to a processor.
The experimental results show that ParallelDHL is comparable to RDFox when handling
the LUBM and the YAGO ontologies.
